% ==========================================================
\section{Integrating Behavioral Programming into UVL}
\label{sec:concept_bp_integration}
% ==========================================================

This section describes, at a conceptual level, how Behavioral Programming can be integrated into the UVL-based language and its available tooling.
The objective is to extend the language and tools so that models can express BP event constraints and BP-specific top-level features while preserving UVL's core philosophy and backward compatibility.

% ----------------------------------------------------------
\subsection{ANTLR grammar extensions: \texttt{uvl-parser}}
\label{subsec:bp-antlr-extensions}
% ----------------------------------------------------------

The \texttt{uvl-parser} repository is maintained by the UVL creators and contains the ANTLR4 grammar files that define UVL's concrete syntax~\cite{sundermann_uvlparser_2023, noauthor_universal-variability-languageuvl-parser_2026}.
ANTLR4 (Another Tool for Language Recognition) is a parser generator that takes a formal grammar as input and produces a parser in a target host language~\cite{parr_definitive_2013, noauthor_antlrantlr4_2026}.
From these grammars, ANTLR4 generates parsers targeting multiple host languages -- most notably Java -- enabling the toolchain to parse \texttt{.uvl} files into parse trees that downstream components consume.
Extending this grammar to support BP introduces two concrete mismatches with the existing rules.

According to the ANTLR grammar, UVL permits a single top-level feature at the model root.
BP models extend this by requiring two additional, named top-level features at the root: \texttt{Env} and \texttt{Config}.
These features are mandatory for BP models, but they do not fit naturally into the ordinary feature hierarchy.
Because the existing grammar enforces a fixed single-feature root, it currently rejects models that include \texttt{Env} or \texttt{Config}.
To support BP while preserving compatibility, the grammar must be extended so these two required top-level features are accepted in a backward-compatible way.
Importantly, this allowance must be limited to the default single top-level feature plus the two BP-specific sections, and must not permit arbitrary additional top-level features.
Standard UVL models that do not include these features should continue to parse unchanged when BP parsing is not enabled.

% [listing - ANTLR rules ?]

The existing constraint alternatives are shaped around classical UVL constraint forms: boolean and mathematical combinations, common expressions and functions (e.g., sum aggregates), and cross-tree relations between features.
Because the base grammar already includes single-reference constraint shapes for these use cases, the same approach can be reused for BP single-reference constraints such as \texttt{requested}, \texttt{waited\_for}, \texttt{blocked}, and \texttt{selected}.
Reusing that approach may require lexical disambiguation for the event-role names so they can appear unambiguously as attribute identifiers.
Adding lexer tokens for those names is acceptable solely to aid parsing, but the parsed result should represent them as ordinary attributes rather than special BP operators.
In contrast, \texttt{conflicting} is a true multi-reference constraint that binds several feature references simultaneously and is not covered by existing single-reference forms.
To support it, an explicit multi-reference production must be added to the grammar.
These BP-specific constraints should be grouped as a distinct alternative, accepted only when BP parsing is enabled, so they do not interfere with the existing constraint forms.
All changes should be implemented as a clearly delimited, opt-in extension so the default UVL grammar remains unchanged unless BP parsing is explicitly enabled.
B-threads and b-events themselves are already representable in the base grammar as attributes of features and do not require bespoke grammar constructs.
Their semantics can be handled by the GLSP server or other tooling after successful parsing.

% ----------------------------------------------------------
\subsection{Java metamodel extensions: \texttt{java-fm-metamodel}}
\label{subsec:java-fm-metamodel-extensions}
% ----------------------------------------------------------

The \texttt{java-fm-featuremodel} hosts the Java source model classes and the parser-to-model factories that translate ANTLR4 parse trees into typed feature model objects \cite{noauthor_universal-variability-languagejava-fm-metamodel_2026}.
It is the central source model for the UVL toolchain: higher-level tools, including the GLSP-based VS Code editor, consume the object graph it produces.
Supporting BP requires some modifications.

% [IMAGE: class diagram showing FeatureModel as the base class and BPFeatureModel as its extension, with the BP constraint class hierarchy alongside]

The metamodel must distinguish cleanly between a standard \texttt{FeatureModel} and a BP-specific \texttt{BPFeatureModel}.
This separation is architecturally significant: it allows the VS Code extension to be deployed as two distinct variants, one targeting plain UVL authoring and one targeting BP-enhanced authoring, each activating only the capabilities relevant to its context.
The \texttt{BPFeatureModel} subclasses the base \texttt{FeatureModel} and is instantiated only when the parser is operating in BP mode.
Consumers that work exclusively with the base class remain entirely unaware of the BP extension and require no modification.

BP introduces constraint semantics that cannot be represented by the existing constraint classes.
An abstract \texttt{BPEventConstraint} base class must be introduced, together with concrete subclasses for each constraint type.
These types correspond directly to the grammar alternatives described in the previous section.
The \texttt{Conflicting} subclass must additionally support multiple references.
Besides the new constraint classes, the \texttt{BPFeatureModel} must also include fields for the \texttt{Env} and \texttt{Config} top-level features.
This should include easy access via the root model object, so that BP-specific tooling can quickly retrieve these features.
The model factory and parser listener must be extended to instantiate these objects when BP parse nodes are encountered, to attach \texttt{Env} and \texttt{Config} features to the \texttt{BPFeatureModel} root, and to resolve event references to their corresponding feature artifacts or imported placeholders during model construction.
The existing unit and integration tests for the \texttt{java-fm-metamodel} must be extended to cover the new BP constructs.