% ==========================================================
\section{Runtime support via FMBP}
\label{sec:concept_fmbp-runtime}
% ==========================================================
 
A central goal of the proposed tooling is to realize the \textit{models@run.time} paradigm~\cite{blair_models_2009, bencomo_modelsruntime_2019}.
In the context of behavioral programs combined with UVL feature models, this means that the graphical representation presented to the user through the GLSP-based editor must not only serve as a static configuration artifact, but must continuously mirror the live state of the executing program.
Conversely, modifications the user applies to the graphical model should propagate to the running system without requiring a restart.
 
FMBP~\cite{felber_comprehensible_2025} provides the foundational layer for this integration.
It is the only existing framework that combines the execution of behavioral programs with a UVL-based configuration mechanism, making it the ideal candidate for realizing the \textit{models@run.time} approach in this context.
The following section describes the conceptual design for extending FMBP to provide the runtime visibility and synchronization features required for a live, interactive model representation in the GLSP editor.

% ----------------------------------------------------------
\subsection{The Runtime State Visibility Problem}
\label{subsec:runtime_state_visibility_problem}
% ----------------------------------------------------------
 
During execution, FMBP maintains all dynamic state internally.
In particular, two categories of runtime information are relevant for a live model representation.

\textbf{Dynamic environment attributes.}
As described in the previous sections, the UVL feature model includes an \texttt{Env} feature whose variables capture measurements of the system's operational context.
While these values are initialized from the model at startup, the running FMBP framework continuously updates them in a local Python object.
The underlying UVL files are never modified during execution; the current values exist exclusively in memory and are invisible to any external observer.

\textbf{Selected B-Events.}
The BP paradigm centers on the synchronization of behavioral threads through the declaration and selection of events.
Whenever the FMBP runtime selects a B-event for execution, this constitutes a meaningful behavioral transition.
Currently, in the examples provided by FMBP, such selections are only reported to the console as log output.
The GLSP server has no means to observe these transitions and therefore cannot reflect them in the graphical model.

Both categories represent live runtime information that the GLSP server requires in order to uphold the \textit{models@run.time} contract.
Since the GLSP server operates as a separate process and has no direct access to the FMBP Python runtime's object space, a communication mechanism must be established that publishes this information proactively.
By maintaining the GLSP server as an independent modeling component, the tool preserves its role as a general-purpose graphical editor that can be extended to multiple runtime systems without tight coupling to FMBP's internals.

% ----------------------------------------------------------
\subsection{Server-Sent Events as a publication mechanism}
\label{subsec:sse_publication_mechanism}
% ----------------------------------------------------------
 
A suitable solution for bridging the gap between the FMBP runtime and the GLSP server is \textit{Server-Sent Events} (SSE).
SSE is a lightweight, unidirectional HTTP-based streaming protocol in which a server pushes a stream of text-encoded events to any connected client~\cite{de_la_torre_using_2019}.
It is well-supported in web environments, requires no custom protocol negotiation, and is straightforwardly implementable in Python~\cite{ramirez_sebastian_server-sent_nodate}.
Crucially, it aligns with the directional nature of the problem: the FMBP runtime is the sole authoritative source of truth for live state, and the GLSP server is a passive consumer that must react to incoming updates.
 
Under this approach, the FMBP runtime exposes an SSE endpoint.
The GLSP server subscribes to this endpoint at startup or when a user initiates a connection.
Then, the server receives a continuous stream of event messages.
Upon receiving an event, the GLSP server applies the encoded state change to the internal model representation and triggers an update of the graphical diagram.
This keeps the visual model synchronized with the executing program without requiring polling or file-based communication.

% ----------------------------------------------------------
\subsection{Extending FMBP: Two Complementary Mechanisms}
\label{subsec:extending_fmbp_concept}
% ----------------------------------------------------------
 
To realize the SSE publication, two architectural extensions to FMBP are introduced.
They are designed to be minimally invasive, preserving full backward compatibility with existing FMBP examples.
 
\paragraph{Decorator for Context Sources.}
The \texttt{Env} attribute problem is addressed by applying the \emph{Decorator} design pattern~\cite{gamma_design_1993} to the existing context source abstraction in FMBP.
FMBP uses a \texttt{ContextSource} that is injected into the \texttt{ContextConfigurationProvider} to supply environment data to the runtime~\cite{felber_comprehensible_2025}.
By wrapping the original context source in a \texttt{ContextSourceStreamDecorator}, every invocation of \texttt{get\_data()} is intercepted.
The decorator first delegates to the original source to obtain the current context snapshot, and then publishes a corresponding SSE to the GLSP server before returning the data to the runtime.
The runtime itself requires no modification; the decorator is inserted at the construction site of the provider, making it trivially opt-in.
 
To avoid redundant network traffic, the decorator performs payload deduplication.
A new event is only emitted when the current snapshot differs from the previously published one.
This prevents the GLSP server from processing unnecessary updates when environment values are stable.
 
\paragraph{BPConfigurator Extension Mechanism.}
The B-event selection problem is addressed by introducing a formal \emph{extension} mechanism into the \texttt{BPConfigurator}, the central orchestration class of FMBP.
\texttt{BPConfigurator} is extended to accept a list of \texttt{BPConfiguratorExtension} objects at construction time.
The extension protocol defines two hooks:
 
\begin{itemize}
    \item \texttt{starting(\ldots)}: invoked once when the behavioral program begins execution.
    \item \texttt{event\_selected(\ldots)}: invoked on every B-event selection, receiving the selected event as an argument.
\end{itemize}
 
A concrete \texttt{BEventStreamerExtension} implements this protocol by publishing an event on each invocation of \texttt{event\_selected}.
As with the decorator, the extension is registered at construction time and does not require any changes to the core runtime logic.

% ----------------------------------------------------------
\subsection{Interaction Flow}
\label{subsec:interaction_flow}
% ----------------------------------------------------------

The combined interaction between FMBP, the SSE layer, and the GLSP server at runtime is illustrated conceptually as follows.
Upon program startup, \texttt{BPConfigurator} invokes the \texttt{starting} hook of all registered extensions, allowing the SSE server to initialize.
The GLSP server connects to the SSE endpoint and begins listening.
During execution, whenever the runtime samples a new context snapshot via the decorated context source, a \textit{context\_update} event is pushed to the GLSP server, which updates the \texttt{Env} variable values shown in the graphical model.
Whenever a b-event is selected, an \textit{event\_selected} server-sent event is dispatched, enabling the GLSP server to highlight the corresponding element in the model.
User edits in the graphical model are written back to the underlying UVL file and are processed by the UVL LSP together with the Z3-based \texttt{ConsistencyChecker}~\cite{felber_comprehensible_2025}.
These updates are subject to runtime restrictions.
The \texttt{ConsistencyChecker} may raise exceptions if model changes diverge from the running state.
Edits to the \texttt{Env} feature's measurement values are not applied to the live runtime.
Runtime-captured context updates must not be persisted back into the UVL files.
These values are transient runtime state and should remain memory-only to avoid conflating observed context with intended configuration.
Only variables explicitly designed for runtime reconfiguration (for example \texttt{Config}) are intended to alter behavior.
 
This bidirectional coupling—runtime-to-model via SSE, and model-to-runtime via the UVL configuration mechanism—constitutes the full realization of the \textit{models@run.time} paradigm in the FMBP setting.