% ==========================================================
\section{GLSP Architecture \& Framework Selection}
\label{sec:concept_glsp}
% ==========================================================

This section explains the architectural choices and framework selections made for implementing the graphical \gls{uvl} editor.
It connects those choices to the thesis goal of providing a \gls{glsp}-based editor extended for context-oriented behavioral programs.
It outlines the separation of responsibilities between server and client components, the role of the graphical model in client-server communication, and the rationale for selecting the server and client technologies.

% ----------------------------------------------------------
\subsection{Server Framework Selection: Java}
\label{subsec:concept_glsp_server}
% ----------------------------------------------------------

\gls{glsp} provides two production-ready server framework implementations~\cite{bork_vision_2024}, one in Java~\cite{noauthor_eclipse-glspglsp-server_2026} and one in TypeScript~\cite{noauthor_eclipse-glspglsp-server-node_2026}.
The Java server framework supports multiple source model strategies.
The most common strategy is to use the \gls{glsp}-native graphical model (GModel) as the primary data structure for both semantic and visual information.
The \gls{glsp} maintainers demonstrate this approach in the workflow example~\cite{noauthor_eclipse-glspglsp-examples_2026}.
An alternative is to integrate an \gls{emf}-based semantic model directly into the server.
The Java framework provides several variants of \gls{emf}-based integration: direct \gls{emf} model integration, \gls{emf} integration paired with a separate notation model, and integration with an EMF.cloud model server for centralized persistence and external semantic authority~\cite{bork_vision_2024}.
The Java server further benefits from a mature ecosystem of multiple examples, reducing implementation friction and providing proven patterns for complex model operations.
The TypeScript server framework, while offering a homogeneous full-stack development experience, lacks the same depth of support for source models.
It supports only a single source model strategy in which the GModel serves as the primary data structure for both semantic and visual information~\cite{noauthor_eclipse-glspglsp-server-node_2026}.

We choose the Java server framework for this work, although it does not adopt the \gls{emf}-based source model.
The Universal Variability Language does not provide an \gls{emf} domain model (an \textit{Ecore} model).
The framework's \gls{emf}-based source model strategies require such an \textit{Ecore} model.
However, the \gls{uvl} maintainers have developed the \texttt{java-fm-metamodel} package~\cite{noauthor_universal-variability-languagejava-fm-metamodel_2026}.
It is a Java library that parses \gls{uvl} files using the \texttt{uvl-parser} and materializes them into structured Java objects suitable for programmatic manipulation~\cite{sundermann_uvlparser_2023}.
This library enables the \gls{glsp} server to load \gls{uvl} models directly, modify them during editing operations, and re-serialize them back to UVL files, providing a stable foundation for model manipulation.
This is an advantage over the TypeScript server, which would require re-implementing the parsing and model-manipulation logic from scratch.
This server-side foundation supports RQ1 by making graphical visualization of \gls{uvl} feature models technically feasible within the editor architecture.

This work adopts a custom dual-model approach in which the server maintains two parallel representations.
The semantic layer consists of \texttt{java-fm-metamodel} objects that represent the \gls{uvl} source model and serve as the authoritative source for model modification.
The visual layer consists of a GModel derived from this semantic representation.
The server is responsible for transforming any source model into a GModel that contains the visual information transmitted to the client~\cite{bork_vision_2024}.
To persist notational information across editing sessions, the server serializes the GModel to JSON, saving node positions, dimensions, and visual groupings in a separate file alongside the \gls{uvl} source file.
When a model is loaded, the server reads both the \gls{uvl} source file and the corresponding GModel file, reconstructing the previous visual representation.
Changes made by the user in the client are translated into semantic modifications on the \texttt{java-fm-metamodel} objects, which are then reflected in the GModel and persisted back to both the \gls{uvl} source file and the GModel JSON file.

Layout metadata, such as element positions and sizes, could potentially be encoded as feature attributes.
The downside of embedding notation into the semantic \texttt{.uvl} file is that it would conflate the two concerns.
This would invalidate the file for any non-\gls{glsp} tooling and include attributes that do not belong in other use cases.

% ----------------------------------------------------------
\subsection{Client Framework Selection: VS~Code}
\label{subsec:concept_glsp_client}
% ----------------------------------------------------------

\gls{glsp} provides platform integration packages for multiple integrated development environments as described in Section~\ref{subsec:glsp_examples}.
We select VS~Code for client deployment for several reasons.
It is the most widely adopted \gls{ide} in contemporary software development practice, reducing adoption barriers for end users~\cite{noauthor_technology_nodate}.
The platform also supports creating and distributing extensions via the \texttt{.vsix} format, enabling a clean separation between the base \gls{uvl} tool and its \gls{bp}-specific extension~\cite{schrepfer_what_nodate}.
Finally, the \gls{glsp} integration for VS~Code is stable, well-documented, and actively maintained, providing a reliable foundation for implementation~\cite{noauthor_eclipse-glspglsp-vscode-integration_2026}.
As this work focuses on establishing a proof of concept, a single environment is sufficient.
Broader \gls{ide} support may be explored in future work.

% ----------------------------------------------------------
\subsection{Feature Division}
\label{subsec:concept_glsp_feature_division}
% ----------------------------------------------------------

A central architectural decision is to separate base \gls{uvl} functionality from \gls{bp}-specific features into two distinct, independently distributed packages.
The base \gls{uvl} tool delivers a standalone \texttt{.vsix} extension providing general-purpose feature modeling capabilities, while the \gls{uvl}-\gls{bp} extension builds upon it by adding behavioral programming semantics.
The separation ensures that users who work exclusively with \gls{uvl} have no dependency on \gls{bp}-specific tooling, and that the \gls{uvl}-\gls{bp} extension can evolve independently without affecting the base editor.
The package structure that realizes this separation is described in the following Section~\ref{subsec:concept_glsp_package_structure}.

\paragraph{Base UVL tool.}
The base tool visualizes the feature hierarchy by presenting feature names, parent-child relations, and constraints in a structured tree view.
It renders group relations such as \textit{alternative} and \textit{or}, and it supports common cross-tree constraints such as \textit{requires} and \textit{excludes}.
Features may carry arbitrary key-value attributes, and the editor also supports feature-level and group-level cardinalities.
Its core editing capabilities cover creating, updating, and deleting features, relations, constraints, and attributes.
The graphical representation adheres to the design requirements outlined in Section~\ref{sec:concept_visualization}.

\paragraph{UVL-BP extension.}
The \gls{uvl}-\gls{bp} extension depends on the base \gls{uvl} tool and introduces additional model elements and runtime capabilities specific to behavioral programming.
It adds two root features, \texttt{Env} and \texttt{Config}, that represent the model's context.
A new b-thread feature type is introduced to represent behavioral threads as first-class model elements.
Different kinds of b-events, namely \textit{requested}, \textit{waited-for}, and \textit{blocked}, may be attached to a b-thread.
\gls{bp}-specific constraints, as defined by Felber and Götz~\cite{felber_behavioral_2025}, are supported via the existing constraint representation of the base tool, thereby extending its semantics without requiring additional model structure.
Finally, the extension realizes a \textit{models@run.time} approach by linking the graphical model to a live behavioral program, reflecting its execution state directly within the editor.
We discuss this integration in detail in the following Sections~\ref{sec:concept_bp_integration} and \ref{sec:concept_fmbp}.
This feature split supports RQ2 by separating the base visualization from \gls{bp} extensions while maintaining an extensible architecture.

% ----------------------------------------------------------
\subsection{Intended Package Structure}
\label{subsec:concept_glsp_package_structure}
% ----------------------------------------------------------

We intend to realize the described separation through a layered package structure that keeps the \gls{uvl} core reusable while isolating \gls{bp}-specific extensions.
On the server side, three main packages are intended.
The \texttt{uvl.metamodel} package contains the adapted parser and metamodel for loading and materializing \gls{uvl} source files.
The \texttt{uvl.glsp} package provides the base \gls{glsp} infrastructure for \gls{uvl} and concentrates all reusable tools, bindings, and shared editor logic.
The \texttt{uvl.bp.glsp} package extends this infrastructure for behavioral programs and adds only the \gls{bp}-specific model elements, constraints, and runtime interactions.
This structure is designed to minimize code duplication by letting the package reuse as much of the common server implementation as possible.

On the client side, we maintain a similar separation.
There are packages for the base \gls{uvl} and for the \gls{bp} extension, each containing the client-side logic relevant to its context.
The VS~Code integration code is shared and not duplicated between both variants.
Instead, profile-specific packaging assembles generated VS~Code extensions from the shared client modules and includes the package set that matches the selected profile.
This allows the repository to produce multiple \texttt{.vsix} artifacts from one codebase while keeping the common integration and rendering code in a single place.
This package structure advances RQ2 by making it possible to add new variants and analyses without duplicating the shared editor foundation.