\TUDoption{abstract}{section}

\begin{abstract}[english,pagestyle=empty.tudheadings]

Behavioral Programming (BP) is a scenario-based paradigm in which programs become increasingly difficult to comprehend at scale, yet existing tooling lacks intuitive graphical support.
We present two Visual Studio Code extensions built on the Eclipse Graphical Language Server Platform (GLSP) that address this gap through graphical management of multiple runtime models for contextual behavioral programs.

The first extension provides a lightweight graphical editor for feature models expressed in the Universal Variability Language (UVL).
The second enriches UVL with BP-specific constructs.
It introduces two new root features, \textit{Env} and \textit{Config}, to capture the contextual aspects of a behavioral program.
To represent BP concepts within the feature model, we model \textit{b-threads} as a dedicated feature type, while \textit{b-events} are expressed as feature attributes directly embedded in the tree structure.
The extension further supports the BP-specific constraints required by this enriched model defined by Felber and Götz.

For runtime monitoring, we adopt the \textit{models@run.time} paradigm by extending Felber's FMBP framework.
During execution, FMBP emits server-sent events for each selected \textit{b-event}, which the GLSP server reflects in the editor via live highlighting.
Dynamic environment attributes are similarly transmitted and rendered on the client side.
Crucially, the toolchain supports simultaneous visualization of multiple runtime models, enabling several independent behavioral programs to be observed and compared within a single graphical environment.

We demonstrate the tools with a water-tank scenario and a multi-drone example featuring four concurrent runtime models.
Both extensions are published on \textit{OpenVSX} with source code available on GitHub.

\end{abstract}