% ==========================================================
\chapter{Conclusion}
\label{chapter:conclusion}
% ==========================================================

In this thesis, we introduce a graphical editor based on the \gls{glsp} framework for \gls{uvl} feature models and extend it to support \gls{bp}-enriched \gls{uvl} feature models.
The toolchain aims to enhance the comprehension of contextual behavioral programs across multiple runtimes. 
In this chapter, we summarize the main contributions of this thesis with regard to the research questions.
Additionally, we outline potential avenues for future work, which could further enhance the capabilities and performance of the developed graphical editor for \gls{uvl} feature models and its integration with behavioral programming concepts.

% ----------------------------------------------------------
\section{Contributions}
\label{sec:conclusion_contributions}
% ----------------------------------------------------------

We present a graphical editor for \gls{uvl} feature models, implemented on top of the \gls{glsp} framework.
The editor supports \gls{bp}-enriched \gls{uvl} feature models and synchronizes model changes with running behavioral programs.
The implementation includes VS~Code extensions that enable users to author and edit \gls{uvl} feature models in the editor.
The toolchain supports \textit{models@run.time} for behavioral programs and allows runtime modifications of feature models to be reflected in the running system.
The repository publishes two separate extensions on \textit{OpenVSX}~\cite{efftinge_eclipse_2020} with one extension targeting base \gls{uvl} models~\cite{ruider_uvl_nodate} and the other targeting \gls{uvl} models enriched with \gls{bp} annotations~\cite{ruider_uvlbp_nodate}.

The contributions of this thesis are relevant to the research questions posed in Section~\ref{sec:intro_goals}.

\begin{description}[leftmargin=1.5cm, labelwidth=1.3cm, labelsep=0.2cm]
    \item[\textbf{RQ1:}]
    The editor demonstrates the feasibility of graphical visualization for \gls{uvl} feature models by providing an interactive visual representation built on \gls{glsp}.
    The editor's design allows users to visually explore the structure and relationships of feature models.
    To improve comprehension of contextual \gls{bp}, we provide visual cues and annotations that help users understand how features relate to behavioral programming concepts.
     
    \item[\textbf{RQ2:}]
    We integrate \gls{bp} concepts into an extensible graphical tool using the \gls{glsp} framework.
    The editor's architecture is designed to separate the core \gls{uvl} modeling concerns from the \gls{bp} extensions, enabling future enhancements.
    It uses a profile-based approach to package two separate VS~Code extensions.
    
    \item[\textbf{RQ3.1:}]
    We extend FMBP to provide runtime information about the executing behavioral program using \gls{sse}.
    The \gls{glsp} server consumes the events and updates the graphical model accordingly.
    This includes transient modifications to the \texttt{Env} attributes and highlights to b-threads when certain b-events are selected in the behavioral program.
    
    \item[\textbf{RQ3.2}]
    We use the \gls{glsp} editor in combination with Felber's FMBP to run behavioral programs.
    Because they operate on the same underlying \gls{uvl} model, user interactions that modify the feature model in the editor can be directly translated into changes in the running behavioral program.
    Primarily, we can modify the constraints and \texttt{Config} attributes of the feature model, thereby altering the runtime behavior of the executed behavioral program.
\end{description}

% ----------------------------------------------------------
\section{Future Work}
\label{sec:conclusion_future_work}
% ----------------------------------------------------------

\textbf{Performance and Scalability.}
Further engineering effort is required to improve the editor's performance for larger and concurrent feature models.
This includes implementing incremental update strategies for the most expensive operations and introducing selective environment modifications that avoid full model refreshes, reducing synchronization overhead and improving perceived responsiveness.

\textbf{Architecture and Maintainability.}
The current \gls{uvl} metamodel and parser would benefit from a refactor to separate grammar concerns from metamodel concerns.
A clearer separation would ease maintenance and enable more modular parsing approaches for future language features.
On the client side, smaller package modularization, similar to BigUML~\cite{metin_developing_2023}, combined with a broader client refactor would simplify reuse and enable deeper integration with the VS~Code platform.

\textbf{Visualization and Interaction.}
The visual presentation could be enhanced by adopting advanced interaction techniques that improve scalability.
Semantic zooming and off-screen representations for less relevant details are promising directions for reducing visual clutter and supporting focus-plus-context interactions~\cite{de_carlo_advanced_2022}.
Interaction feedback should also be strengthened by adding element highlighting and tighter selection feedback; for example, selecting a complex constraint should highlight related features and attributes.
The existing infrastructure already supports highlight dispatching and only requires the corresponding actions and style rules to be completed.

\textbf{Editor Views and Usability.}
Additional editor views would make the tool more flexible for different tasks.
For example, a compact tree view without attributes and a dedicated property panel for constraints would help users who prefer structured or textual representations.

\paragraph{Testing and Evaluation.}
Automated unit and end-to-end tests implemented with Playwright~\cite{noauthor_eclipse-glspglsp-playwright_2026} would increase reliability and enable early detection of regressions.
A user study should also be conducted to assess the editor both as a modeling tool and as a means to improve comprehension of behavioral programming concepts.